---
title: "Analysis of serial dependence in functional time series residuals"
author: "Hassan Maissoro"
output: 
  html_document:
    toc : true
    toc_depth : 4
    number_sections: true
    theme: readable
    highlight: tango
    toc_float:
      collapsed: false
      smooth_scroll: false
    css: style.css
    fig_width: 8
    fig_height: 3
date: "2025-02-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(data.table)
library(ggplot2)
library(ggpubr)
library(magrittr)
library(dygraphs)
library(manipulateWidget)
library(TSA) # For time series tests
library(tseries) # For KPSS test
library(lmtest) # For Durbin-Watson test
library(pracma) # For distance correlation
library(adaptiveFTS)
```

## Introduction
Our aim is to test the assumption that FTS measurement errors are independent.

### Observation Scheme
We consider a stationary FTS $ \{X_n\} $ observed over a **discrete grid** or at **random points**:
$$
Y_{n,i} = X_n(T_{n,i}) + \sigma(T_{n,i})\varepsilon_{n,i}, \quad n = 1, ..., N, \quad i = 1, ..., M_n.
$$
where:
 
- $\{X_n\}$ is a strictly FTS.
- $T_{n,i}$ are i.i.d. samples of a random variable $T$ (or fixed points in **common design**).
- $M_n$ are i.i.d. samples of a random variable $ M$ with expectation $\lambda$.
- $\varepsilon_{n,i}$ are i.i.d. samples of $\varepsilon$ with mean 0 and unit variance.
- $\sigma$ is a Lipschitz continuous function accounting for heteroscedasticity.
- $\{X_n\}, M, T, \varepsilon$ are mutually independent.

### Naradaya-Watson
 
Let $\widehat{X}_n(t;h)$ be the Naradaya-Watson estimator of $X_n$ at $t$ obtained using the sample $\{(T_{n,i}, Y_{n,i}),\ i = 1, ..., M_n\}$ and some bandwidth parameter $h$. The parameter $h$ can be selected by :
 
1. Cross-Validation over each curve;
2. The Adaptive optimal smoothing method proposed by *Golovkine et al. (2022)*. NB : Technical adaptation required for FTS.
3. The adaptive bandwidth optimal bandwidth parameter for mean or (auto)covariance estimation.

### Smoothing residuals
Let the smoothing residuals be for each curve $X_n$:
$$
  \widehat{e}_{n,i} = Y_{n,i} - \widehat{X}_n(T_{n,i};h), \qquad i=1, ..., M_n.
$$
Our aim is to test the serial dependence of the residuals for the whole FTS:

$$
  \{\widehat{e}_{n,i},\; n=1, ..., N,\; i=1, ..., M_n\}.
$$

### Analysis scenarios

1. **Scenario 1** : The Naive approach. We consider $\{\widehat{e}_{n,i},\; n=1, ..., N,\; i=1, ..., M_n\}$ as one sample of FTS.
    - Diagnostic plots : ACF, PACF, etc.
    - Tests in time series analysis : Ljung-Box test, etc.
    - Nonparametric Tests for Serial Dependence : Kendall's tau, Wald–Wolfowitz runs test, Brock-Dechert-Scheinkman (BDS) Test, etc.
  

2. **Scenario 2** : We consider $\{\widehat{e}_{n,i},\; n=1, ..., N,\; i=1, ..., M_n\}$ as $N$ samples $\{\widehat{e}_{n,i},\; i=1, ..., M_n\}$, $n=1, ..., N$.
    - Permutation with respect to $n$ and application of the analysis of **Scenario 1**.
    - Analysis of the serial dependence of each sample $\{\widehat{e}_{n,i},\; i=1, ..., M_n\}$.
    - Analysis of the serial dependence of the whole sample $\{\widehat{e}_{n,i},\; i=1, ..., M_n\}$, $n=1, ..., N$, using
        * Multivariate independence tests Across curves : multivariate Ljung-Box test, etc.
        * Pairwise independence tests
        * Etc.
      

3. **Scenario 3:** We consider $\{\widehat{e}_{n,i},\; n=1, ..., N,\; i=1, ..., M_n\}$ as $N$ samples $\{\widehat{e}_{n,i},\; i=1, ..., M_n\}$, $n=1, ..., N$.
    - For each curve, find a summary statistics of the resiudals. Let $f$ be some summary statistic of the residuals such Variance, Mean, Other Statistics, etc. and define
      $$ \widehat{S}_n = f(\{\widehat{e}_{n,i},\; i=1, ..., M_n\}) $$
    - Study the variations of $\{\widehat{S}_n\}$
    
## Application of the analysis scenarios
 
- We consider the **random** and **Common** resign with sample size $N=150$ and $\lambda = 90$.
- Homoscedasticity noise $\varepsilon \sim \mathcal{N}(0, 0.0625)$.
- We use cross-validation to select the bandwidth parameter $h$ of the Nadaraya-Watson estimator.


```{r eval=FALSE}
# Generate Random design data
dt_rd <- adaptiveFTS::simulate_far(
  N = 150L, lambda = 90L, tdesign = "random",
  Mdistribution = rpois, tdistribution = runif,
  tcommon = seq(0.2, 0.8, len = 50),
  hurst_fun = function(t) hurst_logistic(t, 0.4, 0.8, 50),
  L = 4,
  far_kernel = function(s,t) 9/4 * exp( - (t + 2 * s) ** 2),
  far_mean = function(t) 4 * sin(1.5 * pi * t),
  int_grid = 100L, burnin = 100L,
  remove_burnin = TRUE)
dt_rd <- dt_rd[ttag == "trandom"]
dt_rd[, eps := rnorm(.N, 0, 0.25)]
dt_rd[, Yobs := X + eps]
saveRDS(dt_rd, file = "./data/dt_rd.RDS")

# Generate common design data
dt_cd <- adaptiveFTS::simulate_far(
  N = 150L, lambda = 90L, tdesign = "common",
  Mdistribution = NULL, tdistribution = NULL,
  tcommon = seq(0.2, 0.8, len = 90),
  hurst_fun = function(t) hurst_logistic(t, 0.4, 0.8, 50),
  L = 4,
  far_kernel = function(s,t) 9/4 * exp( - (t + 2 * s) ** 2),
  far_mean = function(t) 4 * sin(1.5 * pi * t),
  int_grid = 100L, burnin = 100L, remove_burnin = TRUE)
dt_cd[, eps := rnorm(.N, 0, 0.25)]
dt_cd[, Yobs := X + eps]
saveRDS(dt_cd, file = "./data/dt_cd.RDS")
```

```{r eval=FALSE}
## NW smoothing : Random design
dt_rd_smooth <- data.table::rbindlist(lapply(1:150, function(id){
  dt_one_curve <- dt_rd[id_curve == id]
  Yobs <- dt_one_curve[, Yobs]
  tobs <- dt_one_curve[, tobs]
  
  n <- length(tobs)
  hcv <- estimate_nw_bw(
    y = Yobs, t = tobs,
    bw_grid = seq(2 / n, n ** (-1/3), len = 100),
    kernel_name = "epanechnikov")
  res_nw <- adaptiveFTS::estimate_nw(y = Yobs, t = tobs, tnew = tobs, h = hcv, kernel_name = "epanechnikov")
  
  dt_one_curve[, cv_opt_bw := res_nw[, h]]
  dt_one_curve[, yhat := res_nw[, yhat]]
  dt_one_curve[, error := Yobs - yhat]
  
  return(dt_one_curve)
}))

saveRDS(dt_rd_smooth, file = "./data/dt_rd_smooth.RDS")

## NW smoothing : Random design
dt_cd_smooth <- data.table::rbindlist(lapply(1:150, function(id){
  dt_one_curve <- dt_cd[id_curve == id]
  Yobs <- dt_one_curve[, Yobs]
  tobs <- dt_one_curve[, tobs]
  
  n <- length(tobs)
  hcv <- estimate_nw_bw(
    y = Yobs, t = tobs,
    bw_grid = seq(2 / n, n ** (-1/3), len = 100),
    kernel_name = "epanechnikov")
  res_nw <- adaptiveFTS::estimate_nw(y = Yobs, t = tobs, tnew = tobs, h = hcv, kernel_name = "epanechnikov")
  
  dt_one_curve[, cv_opt_bw := res_nw[, h]]
  dt_one_curve[, yhat := res_nw[, yhat]]
  dt_one_curve[, error := Yobs - yhat]
  
  return(dt_one_curve)
}))

saveRDS(dt_cd_smooth, file = "./data/dt_cd_smooth.RDS")


```



### Scenario 1

It is the Naive approach. We consider $\{\widehat{e}_{n,i},\; n=1, ..., N,\; i=1, ..., M_n\}$ as one sample of FTS.

  - Diagnostic plots : ACF, PACF, etc.
  - Tests in time series analysis : Ljung-Box test, etc.
  - Nonparametric Tests for Serial Dependence : Kendall's tau, Wald–Wolfowitz runs test, Brock-Dechert-Scheinkman (BDS) Test, etc.

#### Random design
```{r, fig_width = 4, fig_height = 3, fig.align='center'}
dt_rd_smooth <- readRDS("./data/dt_rd_smooth.RDS")
acf(dt_rd_smooth[, error])
pacf(dt_rd_smooth[, error])
```

#### Common design
```{r, fig_width = 4, fig_height = 3, fig.align='center'}
dt_cd_smooth <- readRDS("./data/dt_cd_smooth.RDS")
acf(dt_cd_smooth[, error])
pacf(dt_cd_smooth[, error])
```

    
    



